---
title: "Brief Introduction to Regression, Part 1"
subtitle: "Choosing a Model, Fitting It with OLS, and Reading the Table | Quantitative Histories Workshop"
author: "Ari Kamau (TA)"
date: "08-11-2026"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 2
    number_sections: true
    code_folding: show
    theme: flatly
    highlight: tango
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.align = "center",
  fig.width = 7,
  fig.height = 4.2
)
```

# The plan for tonight


| Section 
|---|---------------------------------------------------------------|
| 1 |**What is regression** 
| 2 |**Choosing a model, and why OLS**
| 3 |**The data: `mtcars`**   
| 4 |**OLS Code: One Predictor**
| 5 |**The Multiple Predictors OLS code** 
| 6 |**The three numbers you check every time**
| 7 |**What can go wrong** 

> **Use this File!!!** Open it in RStudio. Press the green play button  (▶) at the top-right
> of a grey box ("chunk") to run just that chunk. Press **Knit** to turn the whole
> file into a webpage. Nothing here touches your own files. Everything uses
> `mtcars`, a dataset that ships with R.

---

# What is regression?

> "Regression is a statistical tool to estimate the relationship(s) between a
> **dependent variable** ($y$ or outcome variable) and one or more
> **independent variables** ($x$ or predicting variables)."
> Yang (2017), *Encyclopedia of Big Data*


- **Dependent variable** ($y$, outcome, left-hand side): the thing you are trying
  to explain. Tonight: `mpg`.
- **Independent variables** ($x$, predictors, regressors, right-hand side): the
  things you think explain it. Tonight: `wt`, `hp`, `am`.

Regression fits a line through a cloud of points:


$$
\underbrace{y_i}_{\text{what actually happened}} \;=\;
\underbrace{\beta_0 + \beta_1 x_i}_{\text{the line}} \;+\;
\underbrace{\varepsilon_i}_{\text{what the line missed}}
$$

Greek letters ($\beta$, $\varepsilon$) are the **population truth** — unknowable.
Hats ($\hat{\beta}$, $\hat{y}$) are what we **estimate from a sample**. R prints
hats.

 

```{r packages}
if (!requireNamespace("ggplot2", quietly = TRUE)) {
  install.packages("ggplot2")
}
library(ggplot2)
data(mtcars) # 32 cars from a 1974 Motor Trend magazine
```

---



# Choosing a model, and why OLS 

Before you can estimate or regress anything, you have to choose a functional form. That choice is a modelling decision, and it comes from theory. Tonight we choose the simplest one available: a straight line, linear in its parameters.

Specifcally, we will use **Ordinary Least Squares(OLS)**.

## Why OLS: ease of use and ubiquity



**1. Ease of use.** OLS has a closed-form answer. You can compute it by hand:

$$
\hat{\beta}_1 = \frac{\mathrm{Cov}(x,y)}{\mathrm{Var}(x)}
\qquad
\hat{\beta}_0 = \bar{y} - \hat{\beta}_1\bar{x}
$$

Of all the lines you could draw, OLS picks the one that makes the **sum of
squared residuals** as small as possible or the **Line of Best Fit**:

$$
\min_{\hat{\beta}_0,\,\hat{\beta}_1} \sum_{i=1}^{n} e_i^2
= \min \sum_{i=1}^{n} \left(y_i - \hat{y}_i\right)^2
\qquad\text{where } e_i = y_i - \hat{y}_i
$$

Squaring stops positive and negative misses from cancelling, and it is what makes
the closed form possible.

**2. Ubiquity.** 
When you read a paper in economic history, political science, or
sociology, the table is overwhelmingly an OLS style table. Learning to read one will allow you
to start reading the empirical literature in your field.

**Note:** Under a short list of assumptions, OLS is *unbiased*. Its expected value should be the observed relationship, $E[\hat{\beta}] = \beta$. Those assumptions are
the subject of section 7.

---



# The data

```{r mtcars-summary}
head(mtcars)
summary(mtcars[, c("mpg", "wt", "hp", "disp", "am", "cyl")])
```

| Name   | Meaning                        | Type |
|--------|--------------------------------|------|
| `mpg`  | Miles per gallon               | continuous — **our outcome** |
| `wt`   | Weight (1000s of lbs)          | continuous |
| `hp`   | Gross horsepower               | continuous |
| `disp` | Engine displacement (cu. in.)  | continuous |
| `am`   | Transmission: 0 = auto, 1 = manual | **dummy** (0/1) |
| `cyl`  | Number of cylinders (4, 6, 8)  | categorical |

---

# OLS Code: One predictor

## Reading a Graph: One predictor


```{r simple-lm, results = 'hide', echo = TRUE}
reg1 <- lm(mpg ~ wt, data = mtcars)
summary(reg1)
```

```{r plot-ols}
ggplot(mtcars, aes(wt, mpg)) +
  geom_segment(
    aes(xend = wt, yend = fitted(reg1)),
    colour = "#E74C3C",
    alpha = 0.6
  ) + # the residuals
  geom_point(size = 2.6, colour = "#2C3E50") +
  geom_smooth(
    method = "lm",
    se = TRUE,
    colour = "#18BC9C",
    fill = "#18BC9C",
    alpha = 0.15
  ) +
  labs(
    title = "OLS minimizes the SUM of the SQUARES of the red segments",
    subtitle = "Each red segment is a residual: what the line missed",
    x = "Weight (1000 lbs)",
    y = "Miles per gallon"
  ) +
  theme_minimal(base_size = 12)
```

## Reading the output

```

Call:
lm(formula = mpg ~ wt, data = mtcars)

Coefficients:
            Estimate Std. Error t value Pr(>|t|)
(Intercept)  37.2851     1.8776  19.858  < 2e-16 ***
wt           -5.3445     0.5591  -9.559 1.29e-10 ***
```

| Column | Symbol | What it means |
|---|---|---|
| Estimate | $\hat{\beta}_j$ | Change in $y$ for a **one-unit** change in $x_j$ |
| Std. Error | $SE(\hat{\beta}_j)$ | How much $\hat{\beta}_j$ would bounce around across repeated samples |
| t value | $t = \hat{\beta}_j / SE(\hat{\beta}_j)$ | How many standard errors the estimate sits from zero |
| Pr(>&#124;t&#124;) | $p$ | Probability of a $t$ this extreme **if** $\beta_j = 0$ |
| `***` | | $p < 0.001$. `**` < 0.01, `*` < 0.05, `.` < 0.1 |
| Null Hypotheis | | $p < 0.001$. `**` < 0.01, `*` < 0.05, `.` < 0.1 |

**In plain English:** every extra 1,000 lbs costs about 5.34 mpg, and we are very
confident that isn't zero.



## Sum of Squares

Regression is bookkeeping on variation. Total variation in $y$ splits into the
part the model explains and the part it doesn't:

$$
\underbrace{\sum(y_i - \bar{y})^2}_{SS_{\text{tot}} \text{ (Total)}} =
\underbrace{\sum(\hat{y}_i - \bar{y})^2}_{SS_{\text{reg}} \text{ (Explained)}} +
\underbrace{\sum(y_i - \hat{y}_i)^2}_{SS_{\text{res}} \text{ (Residual)}}
$$

| Notation | Name | What it is |
|---|---|---|
| $SS_{\text{tot}}$ | **Total** Sum of Squares | How much $y$ varies in total, around its own mean |
| $SS_{\text{reg}}$ | **Regression** (Explained) Sum of Squares | The part of that variation the model accounts for |
| $SS_{\text{res}}$ | **Residual** Sum of Squares | The part left over — what the line missed |

## Reading a Full table

```{r reg tabl}
reg1 <- lm(mpg ~ wt, data = mtcars)
summary(reg1)
```


| Column | Symbol | What it means |
|---|---|---|
| Residual standard error | $\hat{\sigma} = \sqrt{\dfrac{SS_{\text{res}}}{n-k-1}}$ | The typical miss, in the units of $y$ — here **3.046 mpg**. The square root of the leftover **variance**, i.e. $SS_{\text{res}}$ spread over the degrees of freedom |
| Degrees of freedom | $df = n - k - 1$ | Independent information left after estimating: $32 - 1 - 1 = 30$. Every predictor spends one. Fewer df ⇒ less **precision** and a fatter t-distribution ⇒ you need a bigger $t$ to reach the same $p$ |
| Multiple R-squared | $R^2 = \dfrac{SS_{\text{reg}}}{SS_{\text{tot}}}$ | The share of the total **variance** in $y$ the model accounts for — the sum-of-squares split above, as a single number. **0.7528** here. It never falls when you add a variable |
| Adjusted R-squared | $\bar{R}^2$ | $R^2$ charged a fee per predictor, because each one costs a degree of freedom and therefore **precision**. **0.7446** here. This one *can* fall, which is why it's the number to compare models on |
| F-statistic | $F = \dfrac{SS_{\text{reg}}/k}{SS_{\text{res}}/(n-k-1)}$ | A ratio of **variances**: explained per predictor over unexplained per degree of freedom. The t-tests ask "is *this one* coefficient zero?" — F asks "are *all of them together* zero?" |
| `on 1 and 30 DF` | $k$, $\; n-k-1$ | 1 numerator df (one per predictor) and 30 denominator df — the same df as above |
| `p-value:` (last line) | $p$ for $F$ | The p-value of the **joint** test, not of any single coefficient. Here $1.29\times10^{-10}$ |


## $R^2$: how much did we explain?

$$
R^2 = \frac{SS_{\text{reg}}}{SS_{\text{tot}}} = 1 - \frac{SS_{\text{res}}}{SS_{\text{tot}}}
\qquad
\bar{R}^2 = 1 - \frac{SS_{\text{res}}/(n-k-1)}{SS_{\text{tot}}/(n-1)}
$$

- $R^2$ — the **coefficient of determination** — is the share of total variation
  in $y$ explained by the regression. Ranges 0 to 1.
- $R^2$ **never goes down** when you add a variable, even a column of random
  noise. That's why **adjusted $R^2$** ($\bar{R}^2$) exists: it charges a fee per
  predictor and *can* fall. This is a penalty for overfitting.
- A high $R^2$ doesn't mean the model is true. In history and the social
  sciences, $R^2 = 0.15$ can be a real finding.

---

# The Multiple Predictors OLS code 

One predictor is rarely the whole story. Add more:

$$
\hat{y}_i = \hat{\beta}_0 + \hat{\beta}_1 wt_i + \hat{\beta}_2 hp_i + \hat{\beta}_3 am_i
$$

```{r multiple}
reg2 <- lm(mpg ~ wt + hp + am, data = mtcars)
summary(reg2)
```



## Did the extra predictors earn their keep?

```{r compare-models}
data.frame(
  model = c("mpg ~ wt", "mpg ~ wt + hp + am"),
  r_squared = round(c(summary(reg1)$r.squared, summary(reg2)$r.squared), 4),
  adj_r2 = round(
    c(summary(reg1)$adj.r.squared, summary(reg2)$adj.r.squared),
    4
  ),
  resid_se = round(c(summary(reg1)$sigma, summary(reg2)$sigma), 4),
  df_resid = c(reg1$df.residual, reg2$df.residual)
)
```

Adjusted $R^2$ went **up** and the typical miss went **down**, while we spent two
degrees of freedom to get there. That's the trade: every predictor buys fit and
costs precision. Adjusted $R^2$ is the ledger that charges you for it.

---

# Reading a regression table: the three numbers, every time

Say these out loud. You will read hundreds of these tables, and the same three
questions answer most of them. Repetition is cheap.

| Look at | Where it is | The question it answers | The trap |
|---|---|---|---|
| **p-value** | `Pr(>|t|)`, per coefficient | If this coefficient were truly zero, how often would I see an estimate this far from zero by luck? | It is **not** the probability your theory is right, and **not** the probability the null is true |
| **$R^2$ / adjusted $R^2$** | bottom block | What share of the variation in $y$ did the model account for? | $R^2$ always rises when you add variables. Only **adjusted** $R^2$ can fall. High $R^2$ ≠ true model |
| **Degrees of freedom** | `on 28 degrees of freedom` | How much independent information is left after estimating? $df = n - k - 1$ | Fewer df ⇒ fatter t-distribution ⇒ you need a bigger $t$ for the same p-value. Small samples are punished automatically |


---

# What can go wrong (tomorrow)

Everything above is what happens when the model behaves. However:

> **Linear regression has assumptions, and real data breaks them routinely.**

OLS's unbiasedness and trustworthy standard errors are *conditional*
on a short list of assumptions about the model and the error term. 

| # | Assumption | Loosely |
|---|---|---|
| A1 | **Linearity** | The relationship really is linear in the parameters |
| A2 | **Full rank** | No predictor is a perfect combination of the others |
| A3 | **Exogeneity**, $E[\varepsilon \mid X] = 0$ | The errors are unrelated to the predictors — *the crucial one* |
| A4 | **Homoskedasticity & nonautocorrelation** | Constant error variance, uncorrelated errors |
| A5 | **Data generation** | $X$ comes from a mechanism unrelated to $\varepsilon$ |
| A6 | **Normality** | Errors are normally distributed (convenient, not essential) |

The regression table is only as good as the *weakest*
assumption behind it, and R will print the table with exactly the same confident
asterisks either way. **R will never tell you an assumption failed.** It prints
`***` for a broken model just as happily as for a good one.



> "The linear regression model ... is often only the **departure point for the full analysis**, it
> remains the device used to begin almost all empirical research."
> — Greene, *Econometric Analysis*, §2.1



Thank you!!!

---

# Appendix: to read after class

## Vocabulary

| Term | Working definition |
|---|---|
| **Dependent variable** ($y$) | What you're explaining |
| **Independent variable** ($x$) | What you think explains it |
| **Coefficient** ($\hat{\beta}_j$) | Change in $y$ per one-unit change in $x_j$, holding others fixed |
| **Intercept** ($\hat{\beta}_0$) | Predicted $y$ when all $x = 0$; anchors the line's height |
| **Residual** ($e_i = y_i - \hat{y}_i$) | What the line missed, in the sample |
| **Error term** ($\varepsilon_i$) | The population version of the residual — unobservable |
| **OLS** | The rule: pick the line minimizing $\sum e_i^2$ |
| **$SS_{\text{tot}}$** | Total Sum of Squares — all the variation in $y$ |
| **$SS_{\text{reg}}$** | Regression (Explained) Sum of Squares — the part the model accounts for |
| **$SS_{\text{res}}$** | Residual Sum of Squares — the part it doesn't. $SS_{\text{tot}} = SS_{\text{reg}} + SS_{\text{res}}$ |
| **$R^2$** | $SS_{\text{reg}} / SS_{\text{tot}}$ — share of variation explained |
| **Adjusted $R^2$** | $R^2$ with a penalty per predictor; can fall, can go negative |
| **Standard error** | How much an estimate would bounce across samples |
| **t-statistic** | $\hat{\beta}_j / SE(\hat{\beta}_j)$ — distance from zero, in SEs |
| **p-value** | P(result this extreme \| the null is true). Not P(null is true) |
| **F-statistic** | Are *all* the coefficients jointly zero? |
| **Degrees of freedom** | $n - k - 1$: information left after estimating |
| **Dummy variable** | A 0/1 indicator; its coefficient is a difference in intercepts |
| **Partial effect** | "Holding the other predictors fixed" |
| **Unbiased estimator** | $E[\hat{\beta}] = \beta$ — right on average across repeated samples |


## R commands from tonight

| Command | Does |
|---|---|
| `summary(d)` | five-number summary of each column |
| `lm(y ~ x, data = d)` | fit a linear model |
| `lm(y ~ x1 + x2 + x3, data = d)` | multiple regression |
| `summary(model)` | the full output table |
| `fitted(model)` / `resid(model)` | $\hat{y}$ / $e$ |
| `coef(model)` | the coefficients alone |
| `coef(summary(model))` | the coefficient matrix (estimate, SE, t, p) |
| `model$df.residual` | degrees of freedom, $n - k - 1$ |
| `summary(model)$r.squared` / `$adj.r.squared` / `$sigma` | $R^2$ / $\bar{R}^2$ / residual SE |

---

**Sources:** Yang, Q. (2017). "Regression." In L.A. Schintler & C.L. McNeely
(eds.), *Encyclopedia of Big Data*. Springer.
DOI 10.1007/978-3-319-32001-4_174-1.
Greene, W.H. (2003). *Econometric Analysis*, 5th ed. Prentice Hall — Ch. 2 (the
classical linear regression model and its assumptions) and Ch. 3 (least squares,
the sum-of-squares decomposition, and $R^2$).
Data: Henderson & Velleman (1981), *Motor Trend* 1974, via R's `mtcars`.
